---
title: "Using the Group Size Covariate to Correct Size Bias"
description: |
  Overview of the options for specifying a distribution for groups sizes.
author:
  - name: Jackson D. Nowacek, Eric Rexstad, Laura Marshall, Len Thomas
    url: http://distancesampling.org
    affiliation: University of St Andrews
    affiliation_url: https://creem.st-andrews.ac.uk
date: "`r format(Sys.time(), '%B %Y')`"
output: 
  bookdown::html_document2:
    number_sections: false
    toc: true
    toc_depth: 2
    base_format: rmarkdown::html_vignette
pkgdown:
  as_is: true   
bibliography: Group-Size.bib
csl: apa.csl
vignette: >
  %\VignetteIndexEntry{Group Size Covariate}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

## Background

```{r include=FALSE}
knitr::opts_chunk$set(eval=TRUE, echo=TRUE, message=FALSE, warnings=FALSE)
```

```{r loadpack, warning=FALSE, message=FALSE, include=F}
library(knitr)
library(dsims)
library(Distance)
library(tidyverse)
library(extraDistr)
library(sf)

set.seed(3)
```

```{r, echo=FALSE,}
myecho <- TRUE
myeval <- TRUE
opts_chunk$set(
  tidy=TRUE     # display NOT code as typed
)
```

When conducting a simulation study of abundance using distance sampling, the make.population.description() function allows for the specification of individual-level covariates. One of those possible covariates is group size, for when the study includes detection of groups rather than individuals. You can include covariates as either a discrete distribution in a dataframe, or as one of the included distributions and their parameters. The current options for user-input distributions for all covariates, including group size, are:

- Normal

- Log Normal

- Poisson

- Zero-truncated Poisson

- Zero-Truncated Negative Binomial

- Zero-Truncated Poisson Log Normal

Since groups of animals are inherently non-zero in size, the zero-inflated distributions are more natural for simulating group sizes.

Selecting between those distributions is an important consideration, as the mean-variance relationship of the distribution selected should reflect that of your group-size data. The most flexible option is the poisson log normal, where the only restrictions are that the group size mean must be larger than 1, and the variance must be greater than the mean. For the negative binomial, the variance of the group sizes must be between 1.2 and 1.8 times the mean, and the mean must be larger than 3, and for the Poisson, the mean and the variance need to be equal, with a mean of at least 1.

The following simulation study will demonstrate how to implement the group size covariate, and why it is important.

## Simulation Study

Let us consider a population of animals which are typically found in groups of about 10:

```{r}
true_sizes <- data.frame(sizes = rpois(1000, lambda = 10))

# true_sd <- sd(true_sizes$sizes)

ggplot(true_sizes, aes(x = sizes)) +
  geom_histogram(binwidth = 2, fill = "grey", color = "black") +
  theme_bw() +
  labs(x = "True Distribution of Group Sizes", y = "Frequency")
```

We will now create a population with this distribution of group size, with a true number of groups of 400; hence true number of individuals in the population is around 400 times our mean group size of 10, so 4000.



```{r}
# Region set-up
simsquare <- st_bbox(c(xmin = 0, xmax = 2000, ymin = 0, ymax = 1000)) |> 
  st_as_sfc() |> 
  st_as_sf()


region <- make.region(region.name = "Study Area",
                      shape = simsquare,
                      units = "m")

density <- make.density(region = region, x.space = 50, constant = 1)

# Apply mild, asymmetrically distributed hot and low spots
density <- add.hotspot(density, centre = c(500, 800), sigma = 300, amplitude = 0.6)
density <- add.hotspot(density, centre = c(1600, 300), sigma = 400, amplitude = 0.4)
density <- add.hotspot(density, centre = c(1200, 700), sigma = 200, amplitude = -0.5)

# plot(density, region, scale = 1)

density.data <- density@density.surface[[1]]

# Smooth the true density surface
fit.gam <- mgcv::gam(density ~ s(x,y), data = density.data, family = gaussian(link="log"))

gam.density <- make.density(region = region,
                                 x.space = 50,
                                 fitted.model = fit.gam)

plot(gam.density, region, scale = 1)
```

```{r}
covariate.list <- list()

covariate.list$size <- list(list(distribution = "ztruncpois", mean = 10))


# Assign N = 400 groups and simulate group size covariate (~10) via zero-truncated poisson
pop.desc <- 
  make.population.description(region = region,
                              density = gam.density,
                              covariates = covariate.list,
                              N = 400,
                              fixed.N = TRUE)

detect <- make.detectability(key.function = "hn",
                                  scale.param = 30,
                                  truncation = 80)

pop <- generate.population(object = pop.desc, 
                                detectability = detect, 
                                region = region)

plot(pop, region)
```

```{r}
cov.params <- list(size = c(0.10))

detect <- make.detectability(scale.param = 10, 
                             cov.param = cov.params, 
                             truncation = 80)

plot(detect, pop.desc)
```

```{r}
my.pop <- generate.population(pop.desc, detect, region)

transects <- make.design(region = region,
                         transect.type = "line",
                         design = "systematic",
                         truncation = 80)

simulation <- make.simulation()

detmodel.size <- make.ds.analysis(dfmodel = list(~size),
                                  key = "hn",
                                  criteria = "AIC",
                                  truncation = 80)

nsim <- 50

size.cov <- make.simulation(reps = nsim, 
                            design = transects, 
                            pop = pop.desc, 
                            det = detect,
                            ds.analysis = detmodel.size)

size.cov.survey <- run.survey(size.cov)

sim.test <- make.simulation(reps = 1,
                                design = transects,
                                population.description = pop.desc,
                                detectability = detect,
                                ds.analysis = detmodel.size)

# Generate a single instance of a survey: a population, set of transects
# and the resulting distance data
eg.test.survey <- run.survey(sim.test)

# Plot it to view a summary
plot(eg.test.survey, region)
```

```{r}
scatter.smooth(size.cov.survey@dist.data$distance,
               size.cov.survey@dist.data$size,
               xlab="Detection distance",
               ylab="Observed group size",
               main="Detected group size as fn of detection distance")
```

This plot shows that as the detection distance increases, we see fewer groups which we expect, but we also see only larger groups. Since we assume groups are evenly distributed, regardless of distance, we seem to be missing the smaller groups at larger distances. To account for this, we need to include the group size covariate in the detection function.

```{r}
runsim.cov <- run.simulation(size.cov, run.parallel = TRUE)
```

```{r}
hist(runsim.cov@results$expected.size[1,1,1:nsim], 
     main="Computed average group size\nGroup size covariate", 
     xlab="Group size", xlim=c(8,12))
abline(v=unname(covariate.list$size[[1]][[2]][1]), lwd=2, lty=3)
indiv <- runsim.cov@results$individuals$N[1,1:6,nsim+1]
```

Now if we run the simulation without the group size covariate:

```{r}
pop.desc_no_cov <- make.population.description(region = region,
                                        density = gam.density,
                                        N = 400,
                                        fixed.N = TRUE)

detect_no_cov <- make.detectability(key.function = "hn",
                                  scale.param = 30,
                                  truncation = 80)

pop_no_cov <- generate.population(object = pop.desc_no_cov, 
                                detectability = detect_no_cov, 
                                region = region)

transects <- make.design(region = region,
                         transect.type = "line",
                         design = "systematic",
                         truncation = 80)

detmodel.size_no_cov <- make.ds.analysis(dfmodel = ~1, 
                                         key = "hn",
                                         criteria = "AIC",
                                         truncation = 80)

no_size <- make.simulation(reps = nsim, 
                            design = transects, 
                            pop = pop.desc, 
                            det = detect_no_cov,
                            ds.analysis = detmodel.size_no_cov)

survey_no_cov <- run.survey(no_size)

sim.test_no_cov <- make.simulation(reps = 1,
                                design = transects,
                                population.description = pop.desc_no_cov,
                                detectability = detect_no_cov,
                                ds.analysis = detmodel.size_no_cov)

# Generate a single instance of a survey: a population, set of transects
# and the resulting distance data
eg.test_no_cov.survey <- run.survey(sim.test_no_cov)

# Plot it to view a summary
plot(eg.test_no_cov.survey, region)
```

```{r}
scatter.smooth(survey_no_cov@dist.data$distance,
               survey_no_cov@dist.data$size,
               xlab="Detection distance",
               ylab="Observed group size",
               main="Detected group size as fn of detection distance")
```

```{r}
no.size.sim <- run.simulation(no_size, run.parallel = TRUE)
```

First, the necessary syntax to include the group size covariate in the detection function is:

```{r}
hist(no.size.sim@results$expected.size[1,1,1:nsim],
     main="Computed average group size\nNo covariate", 
     xlab="Group size", xlim=c(8,12))
abline(v=unname(covariate.list$size[[1]][[2]][1]), lwd=2, lty=3)
nosize.indiv <- no.size.sim@results$individuals$N[1,1:6,nsim+1]
```
















