---
title: "Using the Group Size Covariate to Correct Size Bias"
subtitle: "Introducing More Flexible Functions for Group Size Simulation"
description: |
  Overview of the options for specifying a distribution for groups sizes.
author:
  - name: Jackson D. Nowacek, Eric Rexstad, Laura Marshall, Len Thomas
    url: http://distancesampling.org
    affiliation: University of St Andrews
    affiliation_url: https://creem.st-andrews.ac.uk
date: "`r format(Sys.time(), '%B %Y')`"
output: 
  bookdown::html_document2:
    number_sections: false
    toc: true
    toc_depth: 2
    base_format: rmarkdown::html_vignette
pkgdown:
  as_is: true   
bibliography: Group-Size.bib
csl: apa.csl
vignette: >
  %\VignetteIndexEntry{Group Size Covariate}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{css, echo=FALSE}
code, pre {
  white-space: pre-wrap !important;
  overflow-x: auto !important;
  word-break: normal !important;
  word-wrap: break-word !important;
}
```

```{r, echo=FALSE, warning=FALSE}
options(width = 80)
knitr::opts_chunk$set(
  eval = TRUE, 
  echo = TRUE, 
  message = FALSE, 
  warning = FALSE,
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 80),
  fig.width = 7,
  fig.height = 5
)

myecho <- TRUE
myeval <- TRUE
```

## Introduction

This vignette demonstrates the importance of accounting for the effects of group size on detectability when using distance sampling to estimate population abundance. Additionally, we use this vignette to demonstrate new covariate distributions which have been included in the `dsims` package version 1.0.7 onwards. These include a zero-truncated negative binomial distribution and a zero-truncated Poisson log-normal distribution, which allow for more flexible simulation of discrete covariate values and are particularly relevant for generating group size values. Specifically, these distributions allow the simulation of over-dispersed count data rather than enforcing the fixed mean-variance relationship of the Poisson distribution, over-dispersion is a recurrent issue with group size values in distance sampling data (@segura_2007; @howe_2019). _LM: I'm not sure either of these papers are talking about over-dispersion in cluster sizes sounds more like over-dispersion due to non independence in Howe 2019. Look for better references from Eric's vignettes._

## Background

```{r loadpack, warning=FALSE, message=FALSE, include=F}
library(knitr)
library(dsims)
library(Distance)
library(tidyverse)
library(extraDistr)
library(sf)

set.seed(3)

nsim <- 1000
```

When conducting a simulation study of population abundance using distance sampling, the `make.population.description()` function allows for the specification of individual-level covariates. One of those possible covariates is group size, for when the study involves the detection of groups rather than individuals. `dsims` allows the distribution associated with the covariates to either be a user-defined discrete distribution, where a `dataframe` specifies the possible values and associated probabilities, or alternatively as one of a number of distributions and their associated parameters. When generating an instance of a population during the simulation `dsims` will randomly generate covariate values for each group/individual based on these distributions. Currently, the covariate distributions implemented within `dsims` include:

- Normal

- Log-Normal

- Poisson

- Zero-truncated Poisson

- Zero-Truncated Negative Binomial

- Zero-Truncated Poisson Log-Normal

This simulation study focuses on the group size covariate. Since groups of animals are inherently non-zero in size, the zero-inflated distributions are the more natural choice for simulating group size values. In addition, group size values tend to be over-dispersed, as described above, so we will look to the newly implemented zero-truncated Poisson log-normal distribution. The only restrictions for this distribution are that the group size mean must be larger than 1, and the variance must be greater than the mean (i.e. over-dispersed). The zero-truncated negative binomial distribution is more restrictive and requires that the variance of the group sizes must be between 1.2 and 1.8 times the mean, and the mean must be larger than 3. The zero-truncated Poisson only requires one parameter to be defined as the mean and the variance are equal, and this value must be equal to or great than 1.

The following simulation study will demonstrate:

1.  How to set up a population description so that group size values are included in the population data.

2.  How to define the detectability of the population so that larger groups are more easily detected at greater distances than smaller groups. 

3.  The bias resulting from ignoring the effects of group size on the probablility of detecting a group at a given distance.

4.  How including group size as a covariate in the detection function model corrects the bias associated with larger groups being more detectable at greater distances than smaller groups. 

## Region and Population Setup

Let us consider a population of animals which have a mean group size of 10 and a standard deviation of 3.5 _(LM: note that 3.5^2 is 12.25 so this case of over-dispersion is very mild... is that based on some real-world scenario? I think it is better just to generate values from the actual distribution used in the simulation here can you update that please?)_. We can use the zero-truncated Poisson log-normal to generate count values for this instance of over-dispersion, these are shown in \@ref(fig:gs_values). 

```{r gs_values, fig.cap="Histogram demonstrating 1000 values randomly generated from a zero-truncated Poisson log-normal distribution with mean=10, sd=3.5."}
# simulates a set of group sizes to mimic the simulation setup
true_sizes <- data.frame(sizes = rpois(1000, lambda = 10))

ggplot(true_sizes, aes(x = sizes)) +
  geom_histogram(binwidth = 2, fill = "grey", color = "black") +
  theme_bw() +
  labs(x = "Group Size", y = "Frequency")
```

We will create a population whose group sizes follows this distribution, with a true number of groups of 400; hence true number of individuals in the population is around 400 times our mean group size of 10, so 4000.

```{r}
# Region set-up
simsquare <- st_bbox(c(xmin = 0, xmax = 2000, ymin = 0, ymax = 1000)) |> 
  st_as_sfc() |> 
  st_as_sf()

# uses dsims function to create the appropriate region object
region <- make.region(region.name = "Study Area",
                      shape = simsquare,
                      units = "m")

# uses dsims function to create a basic uniform density
density <- make.density(region = region, x.space = 50, constant = 1)

# Apply mild, asymmetrically distributed hot and low spots
density <- add.hotspot(density, 
                       centre = c(500, 800), 
                       sigma = 300, 
                       amplitude = 0.6)
density <- add.hotspot(density, 
                       centre = c(1600, 300), 
                       sigma = 400, 
                       amplitude = 0.4)
density <- add.hotspot(density, 
                       centre = c(1200, 700), 
                       sigma = 200, 
                       amplitude = -0.5)

# plot(density, region, scale = 1)

# pulls the density surface object from the density object
density.data <- density@density.surface[[1]]

# Smooth the true density surface 
fit.gam <- mgcv::gam(density ~ s(x,y), 
                     data = density.data, 
                     family = gaussian(link="log"))

# uses dsims function to create a density based on the smoothed surface object
gam.density <- make.density(region = region,
                                 x.space = 50,
                                 fitted.model = fit.gam)

# plot(gam.density, region)
```

The following code creates the necessary dsims object for running this analysis, including making the population description, detectability object, and survey design. 

The following code is repeated for both the zero-truncated Poisson distribution, and the zero-truncated Poisson log normal function to show that the two distributions produce distinct simulation results. 

```{r}
covariate.list.pois <- list()
covariate.list <- list()

# uses the previously standard zero-truncated Poisson distribution...
# function to mimic the display distribution
covariate.list.pois$size <- list(list(distribution = "ztruncpois", 
                                 mean = 10))

# uses the newly implemented zero-truncated Poisson log normal distribution...
# function to mimic the display distribution
covariate.list$size <- list(list(distribution = "ztruncpoislognormal", 
                                 mean = 10, 
                                 sd = 3.5))


# Assign N = 400 groups, simulate group size covariate via 
# zero-truncated Poisson function
pop.desc.pois <- make.population.description(region = region,
                                             density = gam.density,
                                             covariates = covariate.list.pois,
                                             N = 400,
                                             fixed.N = TRUE)

# Assign N = 400 groups, simulate group size covariate via 
# zero-truncated Poisson log normal function
pop.desc <- make.population.description(region = region,
                                        density = gam.density,
                                        covariates = covariate.list,
                                        N = 400,
                                        fixed.N = TRUE)

# Define the covariate effect (larger groups = higher detectability)
cov.params <- list(size = c(0.10))

# We must establish the true detectability with size bias included 
# so the simulation actually generates biased survey data
detect <- make.detectability(scale.param = 10, 
                             cov.param = cov.params, 
                             truncation = 80)

# uses dsims function to generate a population using the Poisson 
pop.pois <- generate.population(object = pop.desc.pois, 
                           detectability = detect, 
                           region = region)

# uses dsims function to generate a population
pop <- generate.population(object = pop.desc, 
                           detectability = detect, 
                           region = region)

#uses dsims function to make a survey design object
design <- make.design(region = region,
                      transect.type = "line",
                      design = "systematic",
                      truncation = 80)
```

## Lack of covariate analysis

Now that we have set up our population to include group sizes, we need to set up an analysis that fails to account for those group sizes and observe the resulting bias.

```{r}
# Create an analysis model that ignores the covariate (dfmodel = ~1)
detmodel.size_no_cov <- make.ds.analysis(dfmodel = ~1, 
                                         key = "hn",
                                         criteria = "AIC",
                                         truncation = 80)

# Build the simulation using the biased detection process...  
# but with the unbiased analysis model 
# uses the zero-truncated Poisson function
no_size.pois <- make.simulation(reps = nsim, 
                           design = design, 
                           pop = pop.desc.pois,
                           det = detect,
                           ds.analysis = detmodel.size_no_cov)

# Build the simulation using the biased detection process...  
# but with the unbiased analysis model 
# uses the zero-truncated Poisson log normal function
no_size <- make.simulation(reps = nsim, 
                           design = design, 
                           pop = pop.desc,
                           det = detect,
                           ds.analysis = detmodel.size_no_cov)

# Run a single survey to check the data
# uses the zero-truncated Poisson function
no_cov_sim.pois <- make.simulation(reps = 1,
                                   design = design,
                                   population.description = pop.desc.pois,
                                   detectability = detect,
                                   ds.analysis = detmodel.size_no_cov)

# Run a single survey to check the data
# uses the zero-truncated Poisson log normal function
no_cov_sim <- make.simulation(reps = 1,
                                   design = design,
                                   population.description = pop.desc,
                                   detectability = detect,
                                   ds.analysis = detmodel.size_no_cov)

# uses dsims run survey function to get survey data
# uses the zero-truncated Poisson function
no_cov_survey.pois <- run.survey(no_cov_sim.pois)

# uses dsims run survey function to get survey data
# uses the zero-truncated Poisson log normal function
no_cov_survey <- run.survey(no_cov_sim)

# Plot it to view a summary
# plot(no_cov_survey, region)
```

This plot shows that as the detection distance increases, we see fewer groups which we expect, but we also see only larger groups. Since we assume groups are evenly distributed, regardless of distance, we seem to be missing the smaller groups at larger distances. To account for this, we should include the group size covariate in the detection function. In this first example, we will not do so to see the issues with that approach.

```{r}
scatter.smooth(no_cov_survey@dist.data$distance,
               no_cov_survey@dist.data$size,
               xlab="Detection distance",
               ylab="Observed group size",
               main="Detected group size as fn of detection distance")

scatter.smooth(no_cov_survey.pois@dist.data$distance,
               no_cov_survey.pois@dist.data$size,
               xlab="Detection distance",
               ylab="Observed group size",
               main="Detected group size as fn of detection distance")
```

```{r, eval=FALSE}
# uses dsims function to run the specified simulations
no.size.simulation.pois <- run.simulation(no_size.pois, run.parallel = TRUE)

# uses dsims function to run the specified simulations
no.size.simulation <- run.simulation(no_size, run.parallel = TRUE)
```

```{r}
# plots the histogram of expected group sizes
hist(no.size.simulation.pois@results$expected.size[1,1,1:nsim],
     main="Computed average group size\nNo covariate \n Zero-truncated Poisson", 
     xlab="Group size", xlim=c(8,12))
abline(v=unname(covariate.list.pois$size[[1]][[2]][1]), lwd=2, lty=3)
nosize.indiv <- no.size.simulation.pois@results$individuals$N[1,1:6,nsim+1]

# plots the histogram of expected group sizes
hist(no.size.simulation@results$expected.size[1,1,1:nsim],
     main="Computed average group size\nNo covariate \nZero-truncated Poisson Log Normal", 
     xlab="Group size", xlim=c(8,12))
abline(v=unname(covariate.list$size[[1]][[2]][1]), lwd=2, lty=3)
nosize.indiv <- no.size.simulation@results$individuals$N[1,1:6,nsim+1]
```

```{r}
hist(no.size.simulation.pois@results$individuals$N,
     main="Simulation abundance estimates \n Zero-truncated Poisson", 
     xlab="Group size", xlim=c(2000, 6000))

hist(no.size.simulation@results$individuals$N,
     main="Simulation abundance estimates\n Zero-truncated Poisson Log Normal", 
     xlab="Group size", xlim=c(2000, 6000))
```


Here we see the expected upward bias in both group size and overall abundance caused by not including the group size covariate. To resolve this, we need to include the group size covariate in the model.


```{r}
# creates the model used for simulations using dsims function which 
# includes the group size covariate
detmodel.size <- make.ds.analysis(dfmodel = list(~size),
                                  key = "hn",
                                  criteria = "AIC",
                                  truncation = 80)

# sets up the simulation using dsims function
size.cov <- make.simulation(reps = nsim, 
                            design = design, 
                            pop = pop.desc, 
                            det = detect,
                            ds.analysis = detmodel.size)

# runs simulated survey to get data
size.cov.survey <- run.survey(size.cov)

# specified simulation using the elements above
sim.test <- make.simulation(reps = 1,
                                design = design,
                                population.description = pop.desc,
                                detectability = detect,
                                ds.analysis = detmodel.size)

# Generate a single instance of a survey: a population, set of transects
# and the resulting distance data
eg.test.survey <- run.survey(sim.test)

# Plot it to view a summary
# plot(eg.test.survey, region)
```

```{r}
scatter.smooth(size.cov.survey@dist.data$distance,
               size.cov.survey@dist.data$size,
               xlab="Detection distance",
               ylab="Observed group size",
               main="Detected group size as fn of detection distance")
```

Here is the same plot from the first run, re-iterating the need for the group size covariate we just added. 

```{r}
# runs the specified simulation
runsim.cov <- run.simulation(size.cov, run.parallel = TRUE)
```

```{r}
hist(runsim.cov@results$expected.size[1,1,1:nsim], 
     main="Computed average group size\nGroup size covariate", 
     xlab="Group size", xlim=c(8,12))
abline(v=unname(covariate.list$size[[1]][[2]][1]), lwd=2, lty=3)
indiv <- runsim.cov@results$individuals$N[1,1:6,nsim+1]
```

```{r}
hist(runsim.cov@results$individuals$N,
     main="Simulation abundance estimates", 
     xlab="Group size", xlim=c(2000, 6000))
```


Here we see that the group size covariate corrected the expected group size and expected abundance from the previous run. 


# Conclusion

Using a group size covariate when modelling survey data in which groups of animals are counted is essential for accuracy in estimation of both group size and overall abundance. Using the correct distribution to simulate your group sizes can play an important role in the accuracy of those simulations. Ensuring that your mean-variance relationship in simulation matches reality is crucial for representative simulation work.

