---
title: "Using the Group Size Covariate to Correct Size Bias"
subtitle: "Introducing More Flexible Functions for Group Size Simulation"
description: |
  Overview of the options for specifying a distribution for groups sizes.
author:
  - name: Jackson D. Nowacek, Eric Rexstad, Laura Marshall, Len Thomas
    url: http://distancesampling.org
    affiliation: University of St Andrews
    affiliation_url: https://creem.st-andrews.ac.uk
date: "`r format(Sys.time(), '%B %Y')`"
output: 
  bookdown::html_document2:
    number_sections: false
    toc: true
    toc_depth: 2
    base_format: rmarkdown::html_vignette
pkgdown:
  as_is: true   
bibliography: Group-Size.bib
csl: apa.csl
vignette: >
  %\VignetteIndexEntry{Group Size Covariate}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{css, echo=FALSE}
code, pre {
  white-space: pre-wrap !important;
  overflow-x: auto !important;
  word-break: normal !important;
  word-wrap: break-word !important;
}
```

```{r, echo=FALSE, warning=FALSE}
options(width = 80)
knitr::opts_chunk$set(
  eval = TRUE, 
  echo = TRUE, 
  message = FALSE, 
  warning = FALSE,
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 80),
  fig.width = 7,
  fig.height = 5
)

myecho <- TRUE
myeval <- TRUE
```

Notes:

- the introduction, background section, and the region and population sections are updated.

- I am framing this in the wrong way. I cant simulate values from the two distributions (ztpois, and ztpln) and say that one is doing well and one isnt, because they will both show the correct results for the populations that I drew from their distributions. ztpois vs ztpln needs to be presented as options for simulating data, options that depend on if the true data are over-dispersed or not.

- UPDATE: I have addressed all of these, I think the current version communicates what we want to say.

# Introduction

This vignette demonstrates the importance of accounting for the effects of group size on detectability when conducting distance sampling analyses. The difference in detectability between large and small groups at varying distances is an important source of bias to address when analyzing organisms that live in groups.

For simulation studies, creating group size distributions that accurately reflect real data trends can be challenging with conventional discrete distributions for a few reasons. First, these distributions often include 0 as a potential value, which group sizes can not, Second, they often have prescribed mean-variance relationships, which can fail to reflect the true mean variance relationship in the data. Over-dispersion is a recurrent issue with group size values in distance sampling data (@goyert2016, @conn2012, @crego2023). The new covariate distributions, which have been added to the `dsims` package from version 1.0.7 onwards, include a zero-truncated negative binomial distribution and a zero-truncated Poisson log-normal distribution, which allow for more flexible mean-variance relationship in the data.

# Background

```{r loadpack, warning=FALSE, message=FALSE}
library(knitr)
library(dsims)
library(Distance)
library(tidyverse)
library(extraDistr)
library(sf)

set.seed(3)

nsim <- 1000
```

When conducting a simulation study using distance sampling, the `make.population.description()` function allows for the specification of individual-level covariates. One of those possible covariates is group size, used when the study involves the detection of groups rather than individuals. `dsims` allows the distribution associated with the covariates to be either a user-defined discrete distribution, where dataframe objects specify the possible values and associated probabilities, or one of several included distributions and their associated parameters. When generating an instance of a population during the simulation, `dsims` will randomly generate covariate values for each group/individual based on these distributions. Currently, the covariate distributions implemented within `dsims` include:

- Normal

- Log-Normal

- Poisson

- Zero-truncated Poisson

- Zero-Truncated Negative Binomial

- Zero-Truncated Poisson Log-Normal

This simulation study focuses on the group size covariate. Since groups of animals are inherently non-zero in size, the zero-inflated distributions are the more natural choice for simulating group size values. In addition, group size values tend to be over-dispersed, as described above, so this simulation will show the ability of the newly implemented zero-truncated Poisson log-normal distribution to model over-dispersed group size data. The only restrictions for this distribution are that the mean group size must be larger than 1, and the variance must be greater than the mean (i.e. over-dispersed). The zero-truncated negative binomial distribution is more restrictive and requires that the variance of the group sizes must be between 1.2 and 1.8 times the mean, and the mean must be larger than 3. The zero-truncated Poisson represents the preferred distribution for modelling group size when the distribution is not over-dispersed. This distribution only requires one parameter to be defined as the mean and the variance are equal, and this value must be equal to or great than 1.

The choice between the three zero-truncated distributions will depend on the assumed or estimated mean-variance relationship in the data the user wishes to simulate, not on any other considerations. The characteristics of the distribution of group sizes should point to which distribution is preferable.

# Simulation Study

The following case study will demonstrate:

1.  how to choose between the zero-truncated distributions available based on the data you wish to simulate.

2.  how to use the basic distance sampling functions to simulate a population using a region and population object.

3.  how to specify covariates.

4.  how to set up a population description so that group size values are included in the population data.

5.  how to define the detectability of the population so that larger groups are more easily detected at greater distances than smaller groups.

6.  the bias which results from ignoring the effects of group size on the probability of detecting a group at a given distance.

7.  how including group size as a covariate in the detection function model corrects that bias.

8.  the effectiveness of the zero-truncated poisson log normal distribution for modelling over-dispersed group size data as compared to the zero-truncated poisson distribution.

## Selection of a Group Size Distribution

Species vary significantly in their distribution of group sizes. Those values can also vary within species by season, food availability, or other external factors. To correctly simulate a distribution of group sizes, the right distribution needs to be selected. Let us consider a few examples:

If your study species is a large mammal that typically lives in family groups (lions, elephants, wolves, etc), your distribution of group sizes might look something like this:

```{r}
eg1 <- data.frame(x = rbbinom(n = 10000, size = 100, alpha = 2, beta = 15))

eg1 <- eg1 |>
  filter(x > 0)

eg1 |>
  ggplot(aes(x = x)) + 
  geom_histogram(binwidth = 2) + 
  labs(x = "Group Size", y = "Frequency") +
  theme_bw()

sum_tab <- eg1 |>
  summarise(Mean = mean(x),
            Variance = var(x))

sum_tab |>
  kable(digits = 1)
```

Here, distributions like the truncated poisson or truncated negative binomial would be poor choices, as the data is too over-dispersed to be represented well by those distributions. Here, the zero-truncated poisson log-normal distribution would be the best choice.

In the rare case where group sizes are not over-dispersed or under-dispersed, like the case study of orangutan groups found in @cohen1975, the group size distribution may look like this:

```{r}
eg2 <- data.frame(x = rtpois(n = 100000, lambda = 20))

# eg2 <- eg2 |>
#   filter(x > 0)

eg2 |>
  ggplot(aes(x = x)) + 
  geom_histogram(binwidth = 2) +  
  labs(x = "Group Size", y = "Frequency") +
  theme_bw()

sum_tab <- eg2 |>
  summarise(Mean = mean(x),
            Variance = var(x))

sum_tab |>
  kable(digits = 1)
```

In this case, and in those where the data are even less variable in relation to the mean, using a truncated poisson or normal distribution respectively would be appropriate.

## Region and Population Setup

Distance sampling simulation studies necessitate both a study region and a population. The following code shows how one might create a region from scratch, region objects may also be brought in using a `.shp.` file or an `sf` object. A density surface representing the population abundance in different areas of the region is then created using hotspots. These peaks in the baseline density can be controlled in their position, intensity, and spread.

```{r}
# Region set-up
simsquare <- st_bbox(c(xmin = 0, xmax = 2000, ymin = 0, ymax = 1000)) |> 
  st_as_sfc() |> 
  st_as_sf()

# uses dsims function to create the appropriate region object
region <- make.region(region.name = "Study Area",
                      shape = simsquare,
                      units = "m")

# uses dsims function to create a basic uniform density
density <- make.density(region = region, x.space = 50, constant = 1)

# Apply mild, asymmetrically distributed hot and low spots
density <- add.hotspot(density, 
                       centre = c(500, 800), 
                       sigma = 300, 
                       amplitude = 0.6)
density <- add.hotspot(density, 
                       centre = c(1600, 300), 
                       sigma = 400, 
                       amplitude = 0.4)
density <- add.hotspot(density, 
                       centre = c(1200, 700), 
                       sigma = 200, 
                       amplitude = -0.5)

# plot(density, region, scale = 1)

# pulls the density surface object from the density object
density.data <- density@density.surface[[1]]

# Smooth the true density surface 
fit.gam <- mgcv::gam(density ~ s(x,y), 
                     data = density.data, 
                     family = gaussian(link="log"))

# uses dsims function to create a density based on the smoothed surface object
gam.density <- make.density(region = region,
                                 x.space = 50,
                                 fitted.model = fit.gam)

plot(gam.density, region)
```

## Covariates and Detectability

Now that the region is constructed, we can consider covariates, the population description object, detectability, the population object, and the survey design.

First we specify within our list of covariates one for group size. We will create two populations: one using the zero-truncated poisson log-normal distribution, and one using the zero-truncated poisson distribution. Each of which needs to be specified as an element of a list of covariates.

Next we use those covariate lists, as well as the region and density objects from above, to create population description objects. These include a specification for the number of animals in the study area.

A crucial step in simulation studies where group size is important is the specification of how much more visible at increasing distances larger groups are than smaller groups, as this is the source of the bias. The inclusion of the `cov.param` argument in the `make.detectability` function specifies this process to the detectability function.

Population and survey objects are then created using population description, region, and detectability objects. The survey object also requires specification of point or line transects, as well as a truncation distance.

```{r}
covariate.list.P <- list()
covariate.list.PLN <- list()

# Zero-truncated Poisson Log-Normal (Over-dispersed)
covariate.list.PLN$size <- list(list(distribution = "ztruncpoislognormal", 
                                 mean = 50, 
                                 sd = 30))

# Zero-truncated Poisson (Not over-dispersed)
covariate.list.P$size <- list(list(distribution = "ztruncpois", 
                                 mean = 50))

# Assign N = 300 groups, simulate group size covariate via ZT-Poisson
pop.desc.P <- make.population.description(region = region,
                                        density = gam.density,
                                        covariates = covariate.list.P,
                                        N = 300,
                                        fixed.N = TRUE)

# Assign N = 300 groups, simulate group size covariate via ZT-Poisson Log-Normal
pop.desc.PLN <- make.population.description(region = region,
                                        density = gam.density,
                                        covariates = covariate.list.PLN,
                                        N = 300,
                                        fixed.N = TRUE)

# Define the covariate effect (larger groups = higher detectability)
cov.params <- list(size = c(0.015))

# We must establish the true detectability with size bias included 
# so the simulation actually generates biased survey data
detect <- make.detectability(scale.param = 10, 
                             cov.param = cov.params, 
                             truncation = 80)

# uses dsims function to generate a population for each distribution
pop.P <- generate.population(object = pop.desc.P, 
                           detectability = detect, 
                           region = region)

pop.PLN <- generate.population(object = pop.desc.PLN, 
                           detectability = detect, 
                           region = region)

#uses dsims function to make a survey design object
design <- make.design(region = region,
                      transect.type = "line",
                      design = "systematic",
                      truncation = 80)
```

## Lack of covariate analysis

Now that we have set up our analysis to include groups of individuals that are variably detectable at different distances, we set up an analysis that fails to account for those group sizes and observe the resulting bias.

```{r}
# Create an analysis model that ignores the covariate (dfmodel = ~1)
detmodel.size_no_cov <- make.ds.analysis(dfmodel = ~1, 
                                         key = "hn",
                                         criteria = "AIC",
                                         truncation = 80)

# Build the simulation using the biased detection process...  
# but with the unbiased analysis model 
no_size.P <- make.simulation(reps = nsim, 
                           design = design, 
                           pop = pop.desc.P,
                           det = detect,
                           ds.analysis = detmodel.size_no_cov)

no_size.PLN <- make.simulation(reps = nsim, 
                           design = design, 
                           pop = pop.desc.PLN,
                           det = detect,
                           ds.analysis = detmodel.size_no_cov)

# Run a single survey to check the data structure
no_cov_survey.P <- run.survey(no_size.P)
no_cov_survey.PLN <- run.survey(no_size.PLN)

# Plot it to view a summary
plot(no_cov_survey.P, region)
plot(no_cov_survey.PLN, region)
```

```{r}
scatter.smooth(no_cov_survey.P@dist.data$distance,
               no_cov_survey.P@dist.data$size,
               xlab="Detection distance",
               ylab="Observed group size",
               main="Detected group size as fn of detection distance (P)")

scatter.smooth(no_cov_survey.PLN@dist.data$distance,
               no_cov_survey.PLN@dist.data$size,
               xlab="Detection distance",
               ylab="Observed group size",
               main="Detected group size as fn of detection distance (PLN)")
```

The plot above shows that as the detection distance increases, we see fewer groups, which we expect, but we also see larger groups on average than we do at smaller distances. We expected this, as we specified in the detectability function that larger groups are more visible than smaller groups.

Since group size is evenly distributed over distance, we will miss the smaller groups at larger distances. To account for this, we should include the group size covariate in the detection function. In this first example, we will not do so to see the issues with that approach.

```{r}
# uses dsims function to run the specified simulations
no.size.simulation.P <- run.simulation(no_size.P, run.parallel = TRUE)
no.size.simulation.PLN <- run.simulation(no_size.PLN, run.parallel = TRUE)
```

```{r}
# plots the histogram of expected group sizes
hist(no.size.simulation.P@results$expected.size[1,1,1:nsim],
     main="Computed average group size\nNo covariate \nZero-truncated Poisson", 
     xlab="Group size", xlim = c(40, 70))
abline(v=unname(covariate.list.P$size[[1]][[2]][1]), lwd=2, lty=3)
nosize.indiv <- no.size.simulation.P@results$individuals$N[1,1:6,nsim+1]

hist(no.size.simulation.PLN@results$expected.size[1,1,1:nsim],
     main="Computed average group size\nNo covariate \nZero-truncated Poisson Log Normal", 
     xlab="Group size", xlim = c(40, 70))
abline(v=unname(covariate.list.PLN$size[[1]][[2]][1]), lwd=2, lty=3)
nosize.indiv <- no.size.simulation.PLN@results$individuals$N[1,1:6,nsim+1]
```

```{r}
# True values established in our population descriptions
true_groups <- 300
true_size <- 50
true_indiv <- true_groups * true_size

# Function to extract mean simulation results and calculate percentage bias
extract_results <- function(sim_obj, model_name, dist_name) {
  est_groups <- mean(sim_obj@results$clusters$N[1, 1, 1:nsim], na.rm = TRUE)
  est_size <- mean(sim_obj@results$expected.size[1, 1, 1:nsim], na.rm = TRUE)
  est_indiv <- mean(sim_obj@results$individuals$N[1, 1, 1:nsim], na.rm = TRUE)
  
  tibble(
    Distribution = dist_name,
    Model = model_name,
    `Est. Groups` = est_groups,
    `Groups Bias (%)` = ((est_groups - true_groups) / true_groups) * 100,
    `Est. Size` = est_size,
    `Size Bias (%)` = ((est_size - true_size) / true_size) * 100,
    `Est. Abundance` = est_indiv,
    `Abun. Bias (%)` = ((est_indiv - true_indiv) / true_indiv) * 100
  )
}

# Bind results into a single summary table
res_table <- bind_rows(
  extract_results(no.size.simulation.P, "No Covariate (~1)", "ZT-Poisson"),
  extract_results(no.size.simulation.PLN, "No Covariate (~1)", "ZT-PLN")
)

# Output formatted table
res_table |>
  kable(digits = 1, 
        caption = "Comparison of Estimation Bias Without Size Covariate")
```

When we simulate group sizes using a standard Poisson distribution with a mean of 50, the group sizes are tightly distributed around the mean, as the variance of 50 means a standard deviation of around 7. This lack of variation in group size means that there is not a large enough difference in detectability at increasing distances between the large and small groups to cause a significant bias in the data. Group sizes in the wild are very unlikely to follow this distribution however, most of them follow a distribution better simulated by the zero-truncated Poisson log-normal distribution, as we discussed before. The large positive bias in the zero-truncated Poisson log-normal analysis shows that difference. 

## Including covariate analysis


```{r}
# creates the model used for simulations which includes the group size covariate
detmodel.size <- make.ds.analysis(dfmodel = list(~size),
                                  key = "hn",
                                  criteria = "AIC",
                                  truncation = 80)

# sets up the corrected simulations
size.cov.P <- make.simulation(reps = nsim, 
                            design = design, 
                            pop = pop.desc.P, 
                            det = detect,
                            ds.analysis = detmodel.size)

size.cov.PLN <- make.simulation(reps = nsim, 
                            design = design, 
                            pop = pop.desc.PLN, 
                            det = detect,
                            ds.analysis = detmodel.size)

# Generate a single instance of a survey to check the data
size.cov.survey.P <- run.survey(size.cov.P)
size.cov.survey.PLN <- run.survey(size.cov.PLN)

# Plot to view a summary
plot(size.cov.survey.P, region)
```

```{r}
# scatter.smooth(size.cov.survey@dist.data$distance,
#                size.cov.survey@dist.data$size,
#                xlab="Detection distance",
#                ylab="Observed group size",
#                main="Detected group size as fn of detection distance")
```

Here is the same plot from the first run, re-iterating the need for the group size covariate we just added.

```{r}
# runs the specified simulation
size.simulation.P <- run.simulation(size.cov.P, run.parallel = TRUE)
size.simulation.PLN <- run.simulation(size.cov.PLN, run.parallel = TRUE)
```

```{r}
# plots the histogram of expected group sizes
hist(size.simulation.P@results$expected.size[1,1,1:nsim],
     main="Computed average group size\nIncluding covariate \nZero-truncated Poisson", 
     xlab="Group size", xlim = c(40, 70))
abline(v=unname(covariate.list.P$size[[1]][[2]][1]), lwd=2, lty=3)
nosize.indiv <- size.simulation.P@results$individuals$N[1,1:6,nsim+1]

hist(size.simulation.PLN@results$expected.size[1,1,1:nsim],
     main="Computed average group size\nIncluding covariate \nZero-truncated Poisson Log Normal", 
     xlab="Group size", xlim = c(40, 70))
abline(v=unname(covariate.list.PLN$size[[1]][[2]][1]), lwd=2, lty=3)
nosize.indiv <- size.simulation.PLN@results$individuals$N[1,1:6,nsim+1]
```

This time when we run the analysis, we see that the estimated group size distribution is centered on the true value, demonstrating the importance of the group size covariate.

```{r}
# True values established in our population descriptions
true_groups <- 300
true_size <- 50
true_indiv <- true_groups * true_size

# Function to extract mean simulation results and calculate percentage bias
extract_results <- function(sim_obj, model_name, dist_name) {
  est_groups <- mean(sim_obj@results$clusters$N[1, 1, 1:nsim], na.rm = TRUE)
  est_size <- mean(sim_obj@results$expected.size[1, 1, 1:nsim], na.rm = TRUE)
  est_indiv <- mean(sim_obj@results$individuals$N[1, 1, 1:nsim], na.rm = TRUE)
  
  tibble(
    Distribution = dist_name,
    Model = model_name,
    `Est. Groups` = est_groups,
    `Groups Bias (%)` = ((est_groups - true_groups) / true_groups) * 100,
    `Est. Size` = est_size,
    `Size Bias (%)` = ((est_size - true_size) / true_size) * 100,
    `Est. Abundance` = est_indiv,
    `Abun. Bias (%)` = ((est_indiv - true_indiv) / true_indiv) * 100
  )
}

# Bind results into a single summary table
res_table <- bind_rows(
  extract_results(no.size.simulation.P, "No Covariate (~1)", "ZT-Poisson"),
  extract_results(no.size.simulation.PLN, "No Covariate (~1)", "ZT-PLN"),
  extract_results(size.simulation.P, "Size Covariate (~size)", "ZT-Poisson"),
  extract_results(size.simulation.PLN, "Size Covariate (~size)", "ZT-PLN")
)

# Output formatted table
res_table |>
  kable(digits = 1, 
        caption = "Comparison of Estimation Bias With and Without Size Covariate")
```

When accounting for the distribution properly in the analysis, by using a group size covariate, the positive bias is largely corrected. The difference is especially noticable in the over-dispersed data, where the abundance bias fell from nearly 10% to just 0.1%.

# Conclusion

Using a group size covariate when modelling survey data in which groups of animals are counted is essential for accuracy in estimation of both group size and overall abundance. Using the correct distribution to simulate your group sizes can play an important role in the accuracy of those simulations. Ensuring that your mean-variance relationship in simulation matches reality is crucial for representative simulation work.

# References

Goyert, H. F., Gardner, B., Sollmann, R., Veit, R. R., Gilbert, A. T., Connelly, E. E., & Williams, K. A. (2016). Predicting the offshore distribution and abundance of marine birds with a hierarchical community distance sampling model. Ecological Applications, 26(6), 1797-1815.

Conn, P. B., Laake, J. L., & Johnson, D. S. (2012). A hierarchical modeling framework for multiple observer transect surveys.

Crego, R. D., Wells, H. B., Connette, G., Stabach, J. A., Soit, N., & Thompson, S. (2024). Monitoring spatiotemporal dynamics of large herbivores across an African rangeland using hierarchical multi‐species distance sampling. *African Journal of Ecology*, *62*(1), e13219.

Joel E. Cohen, The size and demographic composition of social groups of wild orang-utans, Animal Behaviour, Volume 23, Part 3, 1975, Pages 543-550, ISSN 0003-3472.
